Anomaly detection with kernel density estimation on manifolds

### 2022-02-27 paper/paper.Rmd
Application done

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: "Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection. For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: true
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

@@

```{r smartmeter, fig.align = 'center', out.width = "80%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=FALSE}
knitr::include_graphics("figures/smartmeter.png")
```

<!-- 5x2 HDRs plot of vkde and hdr, 5levels, 5 ML methods, kdtrees, eps=1 -->
<!-- index top 20 outliers -->
<!-- compare density ranking of VKDE and HDR,  -->
<!-- scatterplot no colors, table with one row for 5 ML methods -->

<!-- top 1 typical points + top 2 outlier, HDR -->
<!-- top 1 typical points + top 2 outlier, VKDE -->
<!-- raw data plot side by side -->
<!-- density plot in two figures -->

>>>>

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: "Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection. For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

@@

```{r electricityoutliers, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.", echo = FALSE, eval=T}
knitr::include_graphics("figures/kde_2densities_raw_electricity.png")
```

\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities. The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth. As to the anomalies, variable bandwidth tend to capture the unusual electricity demand volume when the usage is very low or high. It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time period in ID 2838. In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happens in certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.

```{r electricityhdrsfixed, fig.align = 'center', out.width = "90%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=T}
knitr::include_graphics("figures/hdr_electricity_hdrbox_3id_7dow.png")
```

```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=T}
knitr::include_graphics("figures/electricity_hdrbox_3id_7dow.png")
```

Further insights could be gained by comparing the quantile region plots of electricity demand against the time of week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
Again the distribution of both typical household in the top panel have shown a repeated period-of-the-week usage pattern, with higher usage during meal time on all seven days of the week and slightly higher usage for weekends. However, this repeated pattern in a week window is clearly for the typical household ID 1321. As for the distributions for outliers, the middle row outliers from variable bandwidth has spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week. The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh. These finding show the difference in finding typical and anomalous households with different bandwidth selection.
